Mental Rotation as Bayesian Quadrature

### 2013-12-02 man/nips-2013-bayesian-optimization.tex
Tone down the introduction and conclusion; other minor edits

\begin{abstract}
  Given a computational resource--for example, the ability to
  visualize an object rotating--how do you best make use of it? We
  explored how mental simulation should be used in the classic
  psychological task of determining if two images depict the same
  object in different orientations. We compared two models on this
  mental rotation task, and found that a model based on an optimal
  experiment design for Bayesian quadrature is objectively more
  accurate and qualitatively more consistent with classic behavioral
  data than a simpler model. We suggest that rational models which
  adaptively exploit available resources are promising in their
  ability to characterize metacognitive processes like mental
  simulation.
\end{abstract}

@@

Consider the images on the left in Fig.~\ref{fig:mental-rotation}. In
each panel, are the two depicted objects identical (except for a
rotation), or distinct? When presented with this mental rotation task,
people default to a strategy in which they visualize one object
rotating until it is congruent with the other
\cite{Shepard1971}. There is strong evidence for such mental
simulation: we can imagine three-dimensional objects in our minds and
manipulate them, to a certain extent, as if they were real
\cite{Kosslyn:2009tj}.  However, the use of mental simulation is
predicated on determining appropriate parameters to give the
simulation, analogous to determining exactly what computation should
be passed to a graphics processor.  In the case of the classic mental
rotation task, we might ask: How do people know which way to rotate
the object?  When should one stop rotating and accept the hypothesis
that the objects are different?

Recent work in cognitive science has shown how problems of allocating
cognitive resources to solving computational problems can be analyzed
using the methods of statistical decision theory
\cite{Lieder:2012wg,Vul:2009wy}. We apply this ``rational
metacognition'' approach to the problem of mental rotation. We
investigate several computational solutions to this task and
qualitatively compare them to human mental rotation performance. In
particular, we argue that performing mental rotation can be framed as
integration over a probability distribution, with the direction of
rotation becoming an optimal experiment design problem (or in machine
learning parlance, an active learning problem). We show that recent
work on methods for Bayesian quadrature
\cite{Diaconis:1988uo,OHagan:1991tx,Osborne:2012tm} provides a way to
solve this problem, outperforming simpler heuristics for determining
the direction and extent of rotation.

@@

We defined a \naive{} model which performs a hill-climbing search over
the similarity function until it reaches a (possibly local)
maximum. Once a maximum as been found, the model computes an estimate
of $Z$ by linearly interpolating between sampled rotations (e.g.,
Fig.~\ref{fig:li}).

@@

We pick the initial direction of rotation which results in the higher
value of $S$. From then on, at each step we compute $\mu_Z$ and
$\tilde{V}$ to estimate a distribution over the likelihood ratio
$\ell$, i.e.  $p(\ell)\approx\frac{1}{p(X_b)}\ \mathcal{N}(Z\ \vert\
\mu_Z, \sigma_z)$.  We choose $h_0$ when $p(\ell < 1)\geq 0.95$, and
$h_1$ when $p(\ell > 1)\geq 0.95$. Until one of these conditions are
met (or the shape has been fully rotated), the model will continue to
compute rotations and update its estimate of $Z$.  The model may also
change direction based on the \textit{expected} posterior variance of
$Z$ given some new sample $a$.\footnote{This is similar to the full
  posterior variance calculation given in \cite{Osborne:2012tm},
  however we compute the variance given only the current mean estimate
  of $a$.} The model will change directions if doing so would lower
this expected variance. Thus, it is able to actively change its
strategy, unlike the hill-climbing procedure.

@@

In this paper, we asked: how do people use mental simulation?  We
investigated the specific case of mental rotation, using rational
analysis to characterize optimal strategies. We found that an
adaptive, BQ model performs best at the classic rotation task
\cite{Shepard1971}, as it is able to actively monitor the confidence
of its estimate and intelligently choose the direction of rotation. A
simpler model based on heuristics performed much worse: it did not
maintain the linear relationship with the true angle of rotation due
to lack of robustness against local maxima or incorrect rotation
direction.  More broadly, the BQ model provides answers to puzzling
questions surrounding the incremental nature of mental rotation: which
way should the object be rotated, and for how long? This model
formalizes these answers in a way that is qualitatively consistent
with human behavior, both in response time linearity
\cite{Shepard1971} and variability \cite{Just1976}. These results
further indicate that this may be another instance in which people do
appropriately use available computational resources to solve the task
at hand.  Rational, adaptive approaches such as this may therefore be
successful when applied to more complex cognitive processes above and
beyond mental rotation.

>>>>

\begin{abstract}
  Given a computational resource--for example, the ability to
  visualize an object rotating--how do you best make use of it? We
  explored how mental simulation should be used in the classic
  psychological task of determining if two images depict the same
  object in different orientations. We compared two models on this
  mental rotation task, and found that a model based on an optimal
  experiment design for Bayesian quadrature is qualitatively more
  consistent with classic behavioral data than a simpler model. We
  suggest that rational models which adaptively exploit available
  resources are promising in their ability to characterize
  metacognitive processes like mental simulation.
\end{abstract}

@@

Consider the images on the left in Fig.~\ref{fig:mental-rotation}. In
each panel, are the two depicted objects identical (except for a
rotation), or distinct? When presented with this mental rotation task,
people default to a strategy in which they visualize one object
rotating until it is congruent with the other
\cite{Shepard1971}. There is strong evidence for such ``mental
simulation'': we can imagine three-dimensional objects in our minds
and manipulate them, to a certain extent, as if they were real
\cite{Kosslyn:2009tj}.  However, the use of mental simulation is
predicated on determining appropriate parameters to give the
simulation, analogous to determining exactly what computation should
be passed to a graphics processor.  In the case of the classic mental
rotation task, we might ask: How do people know which way to rotate
the object?  When should one stop rotating and accept the hypothesis
that the objects are different?

Recent work in cognitive science has shown how the allocation of
cognitive resources to solving computational problems can be analyzed
using the methods of statistical decision theory
\cite{Lieder:2012wg,Vul:2009wy}. We suggest that this ``rational
metacognition'' approach may also be applied to the problem of mental
rotation. Specifically, we hypothesize that mental rotation can be
framed as integration over a probability distribution, with the
direction of rotation becoming an optimal experiment design problem
(or in machine learning parlance, an active learning problem). In an
initial investigation into this hypothesis, we find that recent
methods for Bayesian quadrature
\cite{Diaconis:1988uo,OHagan:1991tx,Osborne:2012tm}, in contrast to a
simpler heuristic model, provide a possible solution to determining
the direction and extent of rotation.

@@

As a lower bound on performance, we defined a \naive{} model which
performs a hill-climbing search over the similarity function until it
reaches a (possibly local) maximum. Once a maximum as been found, the
model computes an estimate of $Z$ by linearly interpolating between
sampled rotations (e.g., Fig.~\ref{fig:li}).

@@

We pick the initial direction of rotation which results in the higher
value of $S$. From then on, at each step we compute $\mu_Z$ and
$\tilde{V}$ to estimate a distribution over the likelihood ratio
$\ell$, i.e.  $p(\ell)\approx\frac{1}{p(X_b)}\ \mathcal{N}(Z\ \vert\
\mu_Z, \sigma_z)$.  We choose $h_0$ when $p(\ell < 1)\geq 0.95$, and
$h_1$ when $p(\ell > 1)\geq 0.95$.\footnote{We chose a threshold of
  0.95 because the standard confidence interval is 95\%. In the
  future, however, this threshold could be fit to human data.} Until
one of these conditions are met (or the shape has been fully rotated),
the model will continue to compute rotations and update its estimate
of $Z$.  Additionally, the model will change direction if doing so
would lower the \textit{expected} posterior variance of $Z$ given some
new sample $a$.\footnote{This is similar to the full posterior
  variance calculation given in \cite{Osborne:2012tm}, however we
  compute the variance given only the current mean estimate of $a$.}
Thus, it is able to actively change its strategy, unlike the
hill-climbing procedure.

@@

In this paper, we asked: how do people perform mental simulation?  We
performed an initial investigation into the specific case of mental
rotation, using rational analysis to characterize optimal strategies.
We demonstrated that the classic mental rotation task
\cite{Shepard1971} presents a non-trivial computational problem and
cannot be solved with a simple, heuristic-based model. In contrast, an
adaptive, Bayesian Quadrature model provides answers to puzzling
questions surrounding the incremental nature of mental rotation: which
way should the object be rotated, and for how long? This model
formalizes these answers in a way that is qualitatively consistent
with human behavior, both in response time linearity
\cite{Shepard1971} and variability \cite{Just1976}. Although this
research is still in its first stages, these initial results support
the idea that mental rotation may be another instance in which people
do appropriately use available computational resources to solve the
task at hand.
